---
title: "capstone project"
output: html_document
---

1.  Dataset preparations

```{r}
library(GEOquery)
library(tidyverse)
library(ComplexHeatmap)
library(factoextra)
library(ggrepel)
library(GGally)
```

1.1 Load GEO data

```{r}
gse <- getGEO("GSE163151")
```

1.2 Extract metadata

```{r}
metadata <- pData(gse$GSE163151_series_matrix.txt.gz)
```

1.3 Check metadata structure

```{r}
dim(metadata)
colnames(metadata)
head(metadata)
```

1.4 Check sample titles

```{r}
head(metadata$title, 20)
```

1.5 Check sample type

```{r}
table(metadata$source_name_ch1)
```

1.6 Find all characteristics columns

```{r}
char_cols <- grep(
  "characteristics_ch1",
  colnames(metadata),
  value = TRUE
)

char_cols
```

1.7 Print unique values in each characteristics column

```{r}
for(col in char_cols){
  cat("\n---", col, "---\n")
  print(unique(metadata[[col]]))
}
```

2.  Metadata preparation

```{r}
metadata_clean <- metadata |>
  transmute(
    geo_accession = geo_accession,
    sample_name = title,
    sample_type = source_name_ch1,

    disease_group = str_remove(
      characteristics_ch1.1,
      "^disease state: "
    ),

    pathogen = str_remove(
      characteristics_ch1.2,
      "^pathogen: "
    ),

    severity = str_remove(
      characteristics_ch1.4,
      "^disease severity: "
    )
  ) |>
  separate(
    sample_name,
    into = c("sample_prefix", "sample_number"),
    sep = "-",
    remove = FALSE
  )

head(metadata_clean)
```

2.1 Check separated sample names

```{r}
head(
  metadata_clean |>
    select(
      sample_name,
      sample_prefix,
      sample_number
    )
)
```

2.2 Create analysis groups

```{r}
metadata_np <- metadata_clean |>
  filter(sample_type == "NP swab") |>
  mutate(
    analysis_group = case_when(
      disease_group == "COVID-19" ~ "COVID-19",
      disease_group == "Viral acute respiratory illness" ~ "Other Virus",
      disease_group == "Donor control" ~ "Healthy",
      disease_group == "Non-viral acute respiratory illness" ~ "Nonviral",
      TRUE ~ "Exclude"
    )
  )

table(metadata_np$analysis_group)

table(
  metadata_np$analysis_group,
  metadata_np$pathogen
)
```

3.  Main comparison metadata

```{r}
metadata_main <- metadata_np |>
  filter(
    analysis_group %in% c("COVID-19", "Other Virus")
  )

table(metadata_main$analysis_group)
```

3.1 Check main comparison metadata

```{r}
head(metadata_main)

table(
  metadata_main$analysis_group,
  metadata_main$pathogen
)
```

4.  Download supplementary files

```{r}
getGEOSuppFiles(
  "GSE163151",
  makeDirectory = TRUE
)
```

4.1 Check downloaded files

```{r}
list.files(
  "GSE163151",
  recursive = TRUE,
  full.names = TRUE
)
```

4.2 Check tar files

```{r}
tar_files <- list.files(
  "GSE163151",
  pattern = "\\.tar$",
  recursive = TRUE,
  full.names = TRUE
)

tar_files
```

4.3 Extract tar files

```{r}
for(f in tar_files){
  untar(
    f,
    exdir = "GSE163151/samples"
  )
}
```

4.4 Find count files

```{r}
count_files <- list.files(
  "GSE163151",
  pattern = "_counts\\.txt\\.gz$",
  recursive = TRUE,
  full.names = TRUE
)

head(count_files, 20)
length(count_files)
```

4.5 Read count files

```{r}
counts_list <- count_files |>
  set_names(
    basename(count_files) |>
      str_remove("^GSM[0-9]+_") |>
      str_remove("_counts\\.txt\\.gz$")
  ) |>
  map(
    ~ read.delim(
      .x,
      header = TRUE,
      row.names = 1,
      check.names = FALSE
    )
  )
```

4.6 Check count files

```{r}
ncols <- map_int(counts_list, ncol)

table(ncols)

names(counts_list)[ncols != 1]
```

4.7 Remove invalid count file

```{r}
counts_list <- counts_list[ncols == 1]

length(counts_list)
```

4.8 Combine count files

```{r}
counts <- do.call(
  cbind,
  counts_list
)

colnames(counts) <- names(counts_list)

dim(counts)
head(counts)
```

4.9 Check sample matching

```{r}
table(
  metadata_main$sample_name %in% colnames(counts)
)
```

4.10 Keep main comparison samples

```{r}
metadata_main <- metadata_main |>
  filter(
    sample_name %in% colnames(counts)
  )

counts_main <- counts[
  ,
  metadata_main$sample_name,
  drop = FALSE
]
```

4.11 Make order of samples match

```{r}
all(
  metadata_main$sample_name ==
    colnames(counts_main)
)
```

4.12 Check final data

```{r}
dim(metadata_main)

dim(counts_main)

head(metadata_main)

head(counts_main)
```

4.13 Check raw counts

```{r}
all(
  counts_main == round(counts_main)
)

all(
  counts_main >= 0
)
```

5.  Filter low-expression genes

```{r}
keep <- rowSums(counts_main >= 10) >= 10

table(keep)

counts_filtered <- counts_main[keep, ]

dim(counts_filtered)
```

5.1 Check library size

```{r}
library_size <- colSums(counts_filtered)

summary(library_size)
```

5.2 Add library size to metadata

```{r}
metadata_main$library_size <- library_size

head(metadata_main)
```

5.3 Plot library size

```{r}
ggplot(
  metadata_main,
  aes(
    x = analysis_group,
    y = library_size
  )
) +
  geom_boxplot() +
  geom_jitter(
    width = 0.2,
    alpha = 0.5
  ) +
  theme_classic() +
  labs(
    x = "Group",
    y = "Library size"
  )
```

5.4 Check number of detected genes per sample

```{r}
detected_genes <- colSums(counts_filtered > 0)

metadata_main$detected_genes <- detected_genes

summary(detected_genes)
```

5.5 Plot detected genes

```{r}
ggplot(
  metadata_main,
  aes(
    x = analysis_group,
    y = detected_genes
  )
) +
  geom_boxplot() +
  geom_jitter(
    width = 0.2,
    alpha = 0.5
  ) +
  theme_classic() +
  labs(
    x = "Group",
    y = "Detected genes"
  )
```

6.  DESeq2 preparation

```{r}
library(DESeq2)
```

6.1 Set row names of metadata

```{r}
rownames(metadata_main) <- metadata_main$sample_name

all(
  rownames(metadata_main) ==
    colnames(counts_filtered)
)
```

6.2 Set analysis group

```{r}
metadata_main$analysis_group <- factor(
  metadata_main$analysis_group,
  levels = c("Other Virus", "COVID-19")
)

table(metadata_main$analysis_group)
```

6.3 Create DESeq2 dataset

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = counts_filtered,
  colData = metadata_main,
  design = ~ analysis_group
)

dds
```

6.4 Normalize counts

```{r}
dds <- estimateSizeFactors(dds)

normalized_counts <- counts(
  dds,
  normalized = TRUE
)

dim(normalized_counts)
head(normalized_counts)
```

6.5 Variance stabilizing transformation

```{r}
vsd <- vst(
  dds,
  blind = TRUE
)

vsd
```

6.6 PCA

```{r}
pca_data <- plotPCA(
  vsd,
  intgroup = "analysis_group",
  returnData = TRUE
)

percent_var <- round(
  100 * attr(pca_data, "percentVar")
)
```

6.7 Plot PCA

```{r}
ggplot(
  pca_data,
  aes(
    x = PC1,
    y = PC2,
    color = analysis_group
  )
) +
  geom_point(
    size = 3,
    alpha = 0.7
  ) +
  xlab(
    paste0(
      "PC1: ",
      percent_var[1],
      "% variance"
    )
  ) +
  ylab(
    paste0(
      "PC2: ",
      percent_var[2],
      "% variance"
    )
  ) +
  theme_classic()
```

6.8 PCA with sample labels

```{r}
ggplot(
  pca_data,
  aes(
    x = PC1,
    y = PC2,
    color = analysis_group,
    label = name
  )
) +
  geom_point(
    size = 3,
    alpha = 0.7
  ) +
  geom_text_repel(
    size = 2,
    max.overlaps = 20
  ) +
  xlab(
    paste0(
      "PC1: ",
      percent_var[1],
      "% variance"
    )
  ) +
  ylab(
    paste0(
      "PC2: ",
      percent_var[2],
      "% variance"
    )
  ) +
  theme_classic()
```

6.9 Check low-depth samples

```{r}
metadata_main |>
  arrange(library_size) |>
  select(
    sample_name,
    analysis_group,
    pathogen,
    library_size,
    detected_genes
  ) |>
  head(20)
```

6.10 Check highest-depth samples

```{r}
metadata_main |>
  arrange(desc(library_size)) |>
  select(
    sample_name,
    analysis_group,
    pathogen,
    library_size,
    detected_genes
  ) |>
  head(20)
```

6.11 Sample correlation

```{r}
sample_cor <- cor(
  assay(vsd)
)

dim(sample_cor)
```

6.12 Sample correlation heatmap

```{r}
Heatmap(
  sample_cor,
  name = "Correlation",
  show_row_names = FALSE,
  show_column_names = FALSE
)
```

6.13 Find most variable genes

```{r}
gene_variance <- apply(
  assay(vsd),
  1,
  var
)

top_genes <- names(
  sort(
    gene_variance,
    decreasing = TRUE
  )
)[1:50]
```

6.14 Heatmap of most variable genes

```{r}
Heatmap(
  assay(vsd)[top_genes, ],
  name = "Expression",
  show_column_names = FALSE,
  cluster_rows = TRUE,
  cluster_columns = TRUE
)
```

6.15 Prepare COVID-19 vs Healthy metadata

```{r}
metadata_covid_healthy <- metadata_np |>
  filter(
    analysis_group %in% c(
      "COVID-19",
      "Healthy"
    )
  )
```

6.16 Check COVID-19 vs Healthy groups

```{r}
table(
  metadata_covid_healthy$analysis_group
)
```

6.17 Keep matching count samples

```{r}
metadata_covid_healthy <- metadata_covid_healthy |>
  filter(
    sample_name %in% colnames(counts)
  )

counts_covid_healthy <- counts[
  ,
  metadata_covid_healthy$sample_name,
  drop = FALSE
]
```

6.18 Check sample order

```{r}
all(
  metadata_covid_healthy$sample_name ==
    colnames(counts_covid_healthy)
)
```

6.19 Filter low-expression genes

```{r}
keep_covid_healthy <- rowSums(
  counts_covid_healthy >= 10
) >= 10

counts_covid_healthy_filtered <-
  counts_covid_healthy[
    keep_covid_healthy,
  ]

dim(
  counts_covid_healthy_filtered
)
```

7.0 Match metadata rownames with count columns

```{r}
rownames(metadata_covid_healthy) <- metadata_covid_healthy$sample_name

all(
  rownames(metadata_covid_healthy) ==
    colnames(counts_covid_healthy_filtered)
)
```

7.1 Create DESeq2 dataset

```{r}
dds_covid_healthy <- DESeqDataSetFromMatrix(
  countData = counts_covid_healthy_filtered,
  colData = metadata_covid_healthy,
  design = ~ analysis_group
)

dds_covid_healthy
```

7.2 Run DESeq2

```{r}
dds_covid_healthy <- DESeq(
  dds_covid_healthy
)
```

7.3 Extract DEG results

```{r}
res_covid_healthy <- results(
  dds_covid_healthy,
  contrast = c(
    "analysis_group",
    "COVID-19",
    "Healthy"
  )
)

res_covid_healthy
```

7.4 Convert results to dataframe

```{r}
deg_covid_healthy <- as.data.frame(
  res_covid_healthy
) |>
  rownames_to_column("gene")

head(deg_covid_healthy)
```

7.5 Remove NA values

```{r}
deg_covid_healthy <- deg_covid_healthy |>
  filter(
    !is.na(padj)
  )

dim(deg_covid_healthy)
```

7.6 Significant DEGs

```{r}
deg_sig <- deg_covid_healthy |>
  filter(
    padj < 0.05,
    abs(log2FoldChange) >= 1
  )

dim(deg_sig)
```

7.7 Count upregulated and downregulated genes

```{r}
deg_sig |>
  mutate(
    direction = case_when(
      log2FoldChange >= 1 ~ "Up",
      log2FoldChange <= -1 ~ "Down"
    )
  ) |>
  dplyr::count(direction)
```

7.8 Top DEGs

```{r}
deg_sig |>
  arrange(padj) |>
  select(
    gene,
    baseMean,
    log2FoldChange,
    pvalue,
    padj
  ) |>
  head(20)
```

7.9 MA plot

```{r}
plotMA(
  res_covid_healthy,
  ylim = c(-5, 5)
)
```

7.10 Prepare volcano plot data

```{r}
deg_covid_healthy <- deg_covid_healthy |>
  mutate(
    significance = case_when(
      padj < 0.05 & log2FoldChange >= 1 ~ "Up",
      padj < 0.05 & log2FoldChange <= -1 ~ "Down",
      TRUE ~ "Not significant"
    )
  )
```

7.11 Volcano plot

```{r}
ggplot(
  deg_covid_healthy,
  aes(
    x = log2FoldChange,
    y = -log10(padj),
    color = significance
  )
) +
  geom_point(
    alpha = 0.6,
    size = 1.5
  ) +
  geom_vline(
    xintercept = c(-1, 1),
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed"
  ) +
  theme_classic() +
  labs(
    x = "log2 Fold Change",
    y = "-log10 adjusted p-value",
    color = "Significance"
  )
```

7.12 Select top DEGs for heatmap

```{r}
top_deg_genes <- deg_sig |>
  arrange(padj) |>
  slice_head(n = 50) |>
  pull(gene)
```

7.13 VST for COVID-19 vs Healthy

```{r}
vsd_covid_healthy <- vst(
  dds_covid_healthy,
  blind = FALSE
)
```

7.14 DEG heatmap

```{r}
Heatmap(
  assay(vsd_covid_healthy)[top_deg_genes, ],
  name = "Expression",
  show_column_names = FALSE,
  cluster_rows = TRUE,
  cluster_columns = TRUE
)
```

7.15 Save DEG results

```{r}
write.csv(
  deg_covid_healthy,
  "COVID19_vs_Healthy_all_DEGs.csv",
  row.names = FALSE
)

write.csv(
  deg_sig,
  "COVID19_vs_Healthy_significant_DEGs.csv",
  row.names = FALSE
)
```

7.16 Save normalized counts

```{r}
normalized_covid_healthy <- counts(
  dds_covid_healthy,
  normalized = TRUE
)

write.csv(
  normalized_covid_healthy,
  "COVID19_vs_Healthy_normalized_counts.csv"
)
```

7.17 Save metadata

```{r}
write.csv(
  metadata_covid_healthy,
  "COVID19_vs_Healthy_metadata.csv",
  row.names = FALSE
)
```

# 8. Additional Differential-Expression Analyses

```{r}
dir.create(
  "results/differential_expression",
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  "results/signatures",
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  "results/enrichment",
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  "results/drug_repurposing",
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  "figures/differential_expression",
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  "figures/enrichment",
  recursive = TRUE,
  showWarnings = FALSE
)
```

8.1 created a reusable DESeq2 function

```{r}
run_deseq_comparison <- function(
  count_matrix,
  metadata,
  group_1,
  group_2,
  comparison_name
) {
  
  # Select samples from the two requested groups
  metadata_sub <- metadata |>
    filter(
      analysis_group %in% c(group_1, group_2)
    )
  
  # Keep samples that have count data
  metadata_sub <- metadata_sub |>
    filter(
      sample_name %in% colnames(count_matrix)
    )
  
  # Set group_2 as the reference group
  metadata_sub$analysis_group <- factor(
    metadata_sub$analysis_group,
    levels = c(group_2, group_1)
  )
  
  # Match count columns to metadata order
  count_sub <- count_matrix[
    ,
    metadata_sub$sample_name,
    drop = FALSE
  ]
  
  rownames(metadata_sub) <- metadata_sub$sample_name
  
  # Verify sample order
  stopifnot(
    identical(
      rownames(metadata_sub),
      colnames(count_sub)
    )
  )
  
  # Ensure counts are integers
  count_sub <- as.matrix(count_sub)
  storage.mode(count_sub) <- "integer"
  
  # Filter low-expression genes
  keep <- rowSums(count_sub >= 10) >= 10
  count_filtered <- count_sub[keep, , drop = FALSE]
  
  # Create DESeq2 object
  dds <- DESeqDataSetFromMatrix(
    countData = count_filtered,
    colData = metadata_sub,
    design = ~ analysis_group
  )
  
  # Run DESeq2
  dds <- DESeq(dds)
  
  # Extract results: group_1 relative to group_2
  res <- results(
    dds,
    contrast = c(
      "analysis_group",
      group_1,
      group_2
    ),
    alpha = 0.05
  )
  
  # Convert to dataframe
  result_table <- as.data.frame(res) |>
    rownames_to_column("gene") |>
    filter(!is.na(padj)) |>
    mutate(
      direction = case_when(
        padj < 0.05 & log2FoldChange >= 1 ~ "Up",
        padj < 0.05 & log2FoldChange <= -1 ~ "Down",
        TRUE ~ "Not significant"
      ),
      comparison = comparison_name
    ) |>
    arrange(padj)
  
  # Significant genes
  significant_table <- result_table |>
    filter(
      padj < 0.05,
      abs(log2FoldChange) >= 1
    )
  
  # Save tables
  write.csv(
    result_table,
    file.path(
      "results/differential_expression",
      paste0(comparison_name, "_all_genes.csv")
    ),
    row.names = FALSE
  )
  
  write.csv(
    significant_table,
    file.path(
      "results/differential_expression",
      paste0(comparison_name, "_significant_genes.csv")
    ),
    row.names = FALSE
  )
  
  # Create volcano plot
  volcano_plot <- ggplot(
    result_table,
    aes(
      x = log2FoldChange,
      y = -log10(padj),
      color = direction
    )
  ) +
    geom_point(
      alpha = 0.6,
      size = 1.3
    ) +
    geom_vline(
      xintercept = c(-1, 1),
      linetype = "dashed"
    ) +
    geom_hline(
      yintercept = -log10(0.05),
      linetype = "dashed"
    ) +
    scale_color_manual(
      values = c(
        "Up" = "#D73027",
        "Down" = "#4575B4",
        "Not significant" = "grey70"
      )
    ) +
    theme_classic() +
    labs(
      title = comparison_name,
      x = paste0(
        "log2 fold change: ",
        group_1,
        " relative to ",
        group_2
      ),
      y = "-log10 adjusted p-value",
      color = NULL
    )
  
  ggsave(
    filename = file.path(
      "figures/differential_expression",
      paste0(comparison_name, "_volcano.png")
    ),
    plot = volcano_plot,
    width = 8,
    height = 6,
    dpi = 300
  )
  
  # Return all important objects
  list(
    dds = dds,
    metadata = metadata_sub,
    counts = count_filtered,
    results = result_table,
    significant = significant_table,
    volcano = volcano_plot
  )
}
```
8.2 run all comparision
```{r}
covid_vs_healthy <- run_deseq_comparison(
  count_matrix = counts,
  metadata = metadata_np,
  group_1 = "COVID-19",
  group_2 = "Healthy",
  comparison_name = "COVID19_vs_Healthy"
)

virus_vs_healthy <- run_deseq_comparison(
  count_matrix = counts,
  metadata = metadata_np,
  group_1 = "Other Virus",
  group_2 = "Healthy",
  comparison_name = "OtherVirus_vs_Healthy"
)

covid_vs_virus <- run_deseq_comparison(
  count_matrix = counts,
  metadata = metadata_np,
  group_1 = "COVID-19",
  group_2 = "Other Virus",
  comparison_name = "COVID19_vs_OtherVirus"
)
```
8.3 check the results
```{r}
comparison_summary <- tibble(
  comparison = c(
    "COVID-19 vs Healthy",
    "Other Virus vs Healthy",
    "COVID-19 vs Other Virus"
  ),
  
  tested_genes = c(
    nrow(covid_vs_healthy$results),
    nrow(virus_vs_healthy$results),
    nrow(covid_vs_virus$results)
  ),
  
  significant_genes = c(
    nrow(covid_vs_healthy$significant),
    nrow(virus_vs_healthy$significant),
    nrow(covid_vs_virus$significant)
  ),
  
  upregulated = c(
    sum(covid_vs_healthy$significant$direction == "Up"),
    sum(virus_vs_healthy$significant$direction == "Up"),
    sum(covid_vs_virus$significant$direction == "Up")
  ),
  
  downregulated = c(
    sum(covid_vs_healthy$significant$direction == "Down"),
    sum(virus_vs_healthy$significant$direction == "Down"),
    sum(covid_vs_virus$significant$direction == "Down")
  )
)

comparison_summary

write.csv(
  comparison_summary,
  "results/differential_expression/comparison_summary.csv",
  row.names = FALSE
)
```

```{r}
covid_vs_healthy$volcano
virus_vs_healthy$volcano
covid_vs_virus$volcano
```
## Save volcano plots

```{r}
dir.create(
  "figures/differential_expression",
  recursive = TRUE,
  showWarnings = FALSE
)
```

9. Shared Viral Responses
9.1 Combine the two disease-versus-healthy comparisons
```{r}
shared_comparison <- covid_vs_healthy$results |>
  select(
    gene,
    covid_log2FC = log2FoldChange,
    covid_padj = padj,
    covid_direction = direction
  ) |>
  inner_join(
    virus_vs_healthy$results |>
      select(
        gene,
        virus_log2FC = log2FoldChange,
        virus_padj = padj,
        virus_direction = direction
      ),
    by = "gene"
  )
```
9.2 Identify shared genes
```{r}
shared_viral_genes <- shared_comparison |>
  filter(
    covid_padj < 0.05,
    virus_padj < 0.05,
    abs(covid_log2FC) >= 1,
    abs(virus_log2FC) >= 1,
    sign(covid_log2FC) == sign(virus_log2FC)
  ) |>
  mutate(
    shared_direction = if_else(
      covid_log2FC > 0,
      "Shared Up",
      "Shared Down"
    )
  ) |>
  arrange(
    shared_direction,
    covid_padj
  )

table(shared_viral_genes$shared_direction)

write.csv(
  shared_viral_genes,
  "results/signatures/shared_viral_response_genes.csv",
  row.names = FALSE
)
```
9.3 Compare effect sizes
```{r}
fold_change_plot <- ggplot(
  shared_comparison,
  aes(
    x = virus_log2FC,
    y = covid_log2FC
  )
) +
  geom_hline(
    yintercept = 0,
    color = "grey70"
  ) +
  geom_vline(
    xintercept = 0,
    color = "grey70"
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed"
  ) +
  geom_point(
    alpha = 0.35,
    size = 1
  ) +
  geom_point(
    data = shared_viral_genes,
    aes(
      x = virus_log2FC,
      y = covid_log2FC,
      color = shared_direction
    ),
    size = 1.5
  ) +
  scale_color_manual(
    values = c(
      "Shared Up" = "#D73027",
      "Shared Down" = "#4575B4"
    )
  ) +
  theme_classic() +
  labs(
    title = "Shared host responses to COVID-19 and other viruses",
    x = "Other Virus vs Healthy log2 fold change",
    y = "COVID-19 vs Healthy log2 fold change",
    color = NULL
  )

fold_change_plot

ggsave(
  "figures/differential_expression/shared_effect_size_plot.png",
  fold_change_plot,
  width = 8,
  height = 7,
  dpi = 300
)
```

10. COVID-19-Associated Genes
```{r}
covid_associated_genes <- covid_vs_healthy$results |>
  select(
    gene,
    covid_healthy_log2FC = log2FoldChange,
    covid_healthy_padj = padj
  ) |>
  inner_join(
    covid_vs_virus$results |>
      select(
        gene,
        covid_virus_log2FC = log2FoldChange,
        covid_virus_padj = padj
      ),
    by = "gene"
  ) |>
  filter(
    covid_healthy_padj < 0.05,
    covid_virus_padj < 0.05,
    abs(covid_healthy_log2FC) >= 1,
    abs(covid_virus_log2FC) >= 1
  ) |>
  mutate(
    direction = if_else(
      covid_healthy_log2FC > 0,
      "Higher in COVID-19",
      "Lower in COVID-19"
    )
  ) |>
  arrange(covid_virus_padj)

table(covid_associated_genes$direction)

write.csv(
  covid_associated_genes,
  "results/signatures/COVID19_associated_genes.csv",
  row.names = FALSE
)
```
11. GO Over-Representation Analysis
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install(
  c(
    "clusterProfiler",
    "org.Hs.eg.db",
    "enrichplot"
  )
)
```
```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
```

11.1 Create an enrichment function
```{r}
run_go_enrichment <- function(
  significant_table,
  result_table,
  comparison_name
) {
  
  background_genes <- unique(result_table$gene)
  
  up_genes <- significant_table |>
    filter(direction == "Up") |>
    pull(gene) |>
    unique()
  
  down_genes <- significant_table |>
    filter(direction == "Down") |>
    pull(gene) |>
    unique()
  
  go_up <- enrichGO(
    gene = up_genes,
    universe = background_genes,
    OrgDb = org.Hs.eg.db,
    keyType = "SYMBOL",
    ont = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.05,
    readable = TRUE
  )
  
  go_down <- enrichGO(
    gene = down_genes,
    universe = background_genes,
    OrgDb = org.Hs.eg.db,
    keyType = "SYMBOL",
    ont = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.05,
    readable = TRUE
  )
  
  write.csv(
    as.data.frame(go_up),
    file.path(
      "results/enrichment",
      paste0(comparison_name, "_GO_up.csv")
    ),
    row.names = FALSE
  )
  
  write.csv(
    as.data.frame(go_down),
    file.path(
      "results/enrichment",
      paste0(comparison_name, "_GO_down.csv")
    ),
    row.names = FALSE
  )
  
  list(
    up = go_up,
    down = go_down
  )
}
```
11.2 Run GO enrichment
```{r}
go_covid_healthy <- run_go_enrichment(
  covid_vs_healthy$significant,
  covid_vs_healthy$results,
  "COVID19_vs_Healthy"
)

go_virus_healthy <- run_go_enrichment(
  virus_vs_healthy$significant,
  virus_vs_healthy$results,
  "OtherVirus_vs_Healthy"
)

go_covid_virus <- run_go_enrichment(
  covid_vs_virus$significant,
  covid_vs_virus$results,
  "COVID19_vs_OtherVirus"
)
```
11.3 Plot enriched pathways
```{r}
dotplot(
  go_covid_healthy$up,
  showCategory = 15
) +
  ggtitle("COVID-19 vs Healthy: Upregulated pathways")

dotplot(
  go_covid_healthy$down,
  showCategory = 15
) +
  ggtitle("COVID-19 vs Healthy: Downregulated pathways")

dotplot(
  go_covid_virus$up,
  showCategory = 15
) +
  ggtitle("COVID-19 vs Other Virus: Upregulated pathways")

dotplot(
  go_covid_virus$down,
  showCategory = 15
) +
  ggtitle("COVID-19 vs Other Virus: Downregulated pathways")
```

12. Additional Downstream Analysis

12.1 Build GO gene sets directly from Section 11 objects

```{r}
extract_enriched_genes <- function(enrich_result) {
  df <- as.data.frame(enrich_result)

  if (nrow(df) == 0 || !"geneID" %in% colnames(df)) {
    return(character(0))
  }

  genes <- unlist(
    strsplit(
      df$geneID[!is.na(df$geneID)],
      split = "/"
    )
  )

  unique(trimws(genes))
}

gene_sets <- list(
  COVID19_vs_Healthy_GO_up = extract_enriched_genes(
    go_covid_healthy$up
  ),
  COVID19_vs_Healthy_GO_down = extract_enriched_genes(
    go_covid_healthy$down
  ),
  COVID19_vs_OtherVirus_GO_up = extract_enriched_genes(
    go_covid_virus$up
  ),
  COVID19_vs_OtherVirus_GO_down = extract_enriched_genes(
    go_covid_virus$down
  ),
  OtherVirus_vs_Healthy_GO_up = extract_enriched_genes(
    go_virus_healthy$up
  ),
  OtherVirus_vs_Healthy_GO_down = extract_enriched_genes(
    go_virus_healthy$down
  )
)

sapply(gene_sets, length)
```

12.2 Summarize GO gene sets and calculate overlaps

```{r}
dir.create(
  "results/gene_sets",
  recursive = TRUE,
  showWarnings = FALSE
)

gene_summary <- data.frame(
  Gene_Set = names(gene_sets),
  Unique_Genes = sapply(gene_sets, length)
)

gene_summary

write.csv(
  gene_summary,
  "results/gene_sets/gene_set_summary.csv",
  row.names = FALSE
)

overlap_matrix <- outer(
  names(gene_sets),
  names(gene_sets),
  Vectorize(function(a, b) {
    length(
      intersect(
        gene_sets[[a]],
        gene_sets[[b]]
      )
    )
  })
)

rownames(overlap_matrix) <- names(gene_sets)
colnames(overlap_matrix) <- names(gene_sets)

overlap_matrix
```

12.3 Jaccard similarity of GO gene sets

```{r}
jaccard_matrix <- matrix(
  0,
  nrow = length(gene_sets),
  ncol = length(gene_sets)
)

rownames(jaccard_matrix) <- names(gene_sets)
colnames(jaccard_matrix) <- names(gene_sets)

for (i in seq_along(gene_sets)) {
  for (j in seq_along(gene_sets)) {

    intersection_n <- length(
      intersect(
        gene_sets[[i]],
        gene_sets[[j]]
      )
    )

    union_n <- length(
      union(
        gene_sets[[i]],
        gene_sets[[j]]
      )
    )

    jaccard_matrix[i, j] <- ifelse(
      union_n == 0,
      NA,
      intersection_n / union_n
    )
  }
}

round(jaccard_matrix, 3)
```

12.4 Prepare log2 fold-change matrix from Section 8 results

```{r}
fc_all <- covid_vs_healthy$results |>
  dplyr::select(
    gene,
    COVID19_vs_Healthy = log2FoldChange
  ) |>
  dplyr::inner_join(
    covid_vs_virus$results |>
      dplyr::select(
        gene,
        COVID19_vs_OtherVirus = log2FoldChange
      ),
    by = "gene"
  ) |>
  dplyr::inner_join(
    virus_vs_healthy$results |>
      dplyr::select(
        gene,
        OtherVirus_vs_Healthy = log2FoldChange
      ),
    by = "gene"
  )

cat(
  "Number of shared tested genes:",
  nrow(fc_all),
  "\n"
)

head(fc_all)
```

12.5 Correlation of gene-expression signatures

```{r}
fc_matrix <- fc_all |>
  dplyr::select(
    COVID19_vs_Healthy,
    COVID19_vs_OtherVirus,
    OtherVirus_vs_Healthy
  )

cor_pearson <- cor(
  fc_matrix,
  method = "pearson",
  use = "pairwise.complete.obs"
)

cor_spearman <- cor(
  fc_matrix,
  method = "spearman",
  use = "pairwise.complete.obs"
)

cat("===== Pearson correlation =====\n")
print(round(cor_pearson, 3))

cat("\n===== Spearman correlation =====\n")
print(round(cor_spearman, 3))
```

12.6 Correlation heatmap

```{r}
cor_df <- as.data.frame(
  as.table(cor_spearman)
)

colnames(cor_df) <- c(
  "Comparison1",
  "Comparison2",
  "Correlation"
)

correlation_heatmap <- ggplot(
  cor_df,
  aes(
    x = Comparison1,
    y = Comparison2,
    fill = Correlation
  )
) +
  geom_tile(
    color = "white"
  ) +
  geom_text(
    aes(
      label = sprintf(
        "%.2f",
        Correlation
      )
    ),
    size = 5
  ) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  coord_fixed() +
  theme_minimal(
    base_size = 12
  ) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    panel.grid = element_blank()
  ) +
  labs(
    title = "Spearman Correlation of Gene Expression Signatures",
    x = NULL,
    y = NULL,
    fill = "Spearman r"
  )

correlation_heatmap
```

12.7 Define significant DEG sets directly from Section 8 results

```{r}
genes_COVID_H <- unique(
  covid_vs_healthy$significant$gene
)

genes_COVID_OV <- unique(
  covid_vs_virus$significant$gene
)

genes_OV_H <- unique(
  virus_vs_healthy$significant$gene
)

cat(
  "Significant COVID19 vs Healthy:",
  length(genes_COVID_H),
  "\n"
)

cat(
  "Significant COVID19 vs OtherVirus:",
  length(genes_COVID_OV),
  "\n"
)

cat(
  "Significant OtherVirus vs Healthy:",
  length(genes_OV_H),
  "\n"
)

shared_all_three <- Reduce(
  intersect,
  list(
    genes_COVID_H,
    genes_COVID_OV,
    genes_OV_H
  )
)

cat(
  "Three-way shared significant DEGs:",
  length(shared_all_three),
  "\n"
)
```

12.8 DEG overlap UpSet plot

```{r}
deg_sets <- list(
  COVID19_vs_Healthy = genes_COVID_H,
  COVID19_vs_OtherVirus = genes_COVID_OV,
  OtherVirus_vs_Healthy = genes_OV_H
)

if (requireNamespace("UpSetR", quietly = TRUE)) {

  upset_data <- UpSetR::fromList(
    deg_sets
  )

  png(
    "results/DEG_overlap_UpSet.png",
    width = 2400,
    height = 1800,
    res = 300
  )

  UpSetR::upset(
    upset_data,
    sets = c(
      "COVID19_vs_Healthy",
      "COVID19_vs_OtherVirus",
      "OtherVirus_vs_Healthy"
    ),
    order.by = "freq",
    mainbar.y.label = "Number of shared significant DEGs",
    sets.x.label = "Number of significant DEGs",
    text.scale = 1.5
  )

  dev.off()

} else {
  message(
    "UpSetR is not installed. DEG sets were created, but the UpSet plot was skipped."
  )
}
```

12.9 GO enrichment of three-way shared significant DEGs

```{r}
ego_three <- enrichGO(
  gene = shared_all_three,
  universe = unique(
    covid_vs_healthy$results$gene
  ),
  OrgDb = org.Hs.eg.db,
  keyType = "SYMBOL",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.05,
  readable = TRUE
)

cat(
  "Significant GO BP terms:",
  nrow(as.data.frame(ego_three)),
  "\n"
)

dotplot(
  ego_three,
  showCategory = 15,
  title = "GO Biological Process: Three-way Shared DEGs"
)
```

12.10 Define COVID-19-specific and OtherVirus-specific significant DEGs

```{r}
covid_specific <- setdiff(
  genes_COVID_H,
  union(
    genes_COVID_OV,
    genes_OV_H
  )
)

other_specific <- setdiff(
  genes_OV_H,
  union(
    genes_COVID_H,
    genes_COVID_OV
  )
)

cat(
  "COVID19-specific significant DEGs:",
  length(covid_specific),
  "\n"
)

cat(
  "OtherVirus-specific significant DEGs:",
  length(other_specific),
  "\n"
)
```

12.11 GO enrichment of COVID-19-specific and OtherVirus-specific DEGs

```{r}
ego_covid_specific <- enrichGO(
  gene = covid_specific,
  universe = unique(
    covid_vs_healthy$results$gene
  ),
  OrgDb = org.Hs.eg.db,
  keyType = "SYMBOL",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.05,
  readable = TRUE
)

ego_other <- enrichGO(
  gene = other_specific,
  universe = unique(
    virus_vs_healthy$results$gene
  ),
  OrgDb = org.Hs.eg.db,
  keyType = "SYMBOL",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.05,
  readable = TRUE
)

cat(
  "COVID19-specific GO BP terms:",
  nrow(as.data.frame(ego_covid_specific)),
  "\n"
)

cat(
  "OtherVirus-specific GO BP terms:",
  nrow(as.data.frame(ego_other)),
  "\n"
)

write.csv(
  as.data.frame(ego_covid_specific),
  "results/enrichment/COVID19_specific_GO_BP.csv",
  row.names = FALSE
)

write.csv(
  as.data.frame(ego_other),
  "results/enrichment/OtherVirus_specific_GO_BP.csv",
  row.names = FALSE
)
```

12.12 Plot COVID-19-specific and OtherVirus-specific GO results

```{r}
p_covid_GO <- dotplot(
  ego_covid_specific,
  showCategory = 10
) +
  ggtitle(
    "GO Biological Process: COVID19-specific DEGs"
  )

p_other_GO <- dotplot(
  ego_other,
  showCategory = 10
) +
  ggtitle(
    "GO Biological Process: OtherVirus-specific DEGs"
  )

p_covid_GO
p_other_GO
```

12.13 Compare COVID-19-specific and OtherVirus-specific GO terms

```{r}
covid_GO <- as.data.frame(
  ego_covid_specific
) |>
  filter(
    !is.na(p.adjust)
  ) |>
  arrange(
    p.adjust
  ) |>
  slice_head(
    n = 10
  ) |>
  mutate(
    Group = "COVID19-specific"
  )

other_GO <- as.data.frame(
  ego_other
) |>
  filter(
    !is.na(p.adjust)
  ) |>
  arrange(
    p.adjust
  ) |>
  slice_head(
    n = 10
  ) |>
  mutate(
    Group = "OtherVirus-specific"
  )

GO_compare <- bind_rows(
  covid_GO,
  other_GO
) |>
  mutate(
    GeneRatio_num = sapply(
      strsplit(
        as.character(GeneRatio),
        "/"
      ),
      function(x) {
        as.numeric(x[1]) /
          as.numeric(x[2])
      }
    ),
    Description_wrap = stringr::str_wrap(
      Description,
      width = 32
    )
  )

head(GO_compare)
```

12.14 Plot GO Biological Process comparison

```{r}
p_GO_compare <- ggplot(
  GO_compare,
  aes(
    x = GeneRatio_num,
    y = reorder(
      Description_wrap,
      GeneRatio_num
    ),
    size = Count,
    color = p.adjust
  )
) +
  geom_point(
    alpha = 0.85
  ) +
  facet_wrap(
    ~ Group,
    nrow = 1,
    scales = "free"
  ) +
  scale_size_continuous(
    name = "Gene Count",
    range = c(3, 10)
  ) +
  scale_color_gradient(
    name = "Adjusted p-value",
    low = "#E64B35",
    high = "#4C78A8"
  ) +
  labs(
    title = "GO Biological Process Comparison",
    x = "Gene Ratio",
    y = NULL
  ) +
  theme_classic(
    base_size = 14
  ) +
  theme(
    plot.title = element_text(
      size = 22,
      face = "bold",
      hjust = 0.5
    ),
    strip.text = element_text(
      size = 14,
      face = "bold"
    ),
    axis.text.y = element_text(
      size = 9
    )
  )

p_GO_compare

ggsave(
  "figures/enrichment/COVID19_vs_OtherVirus_GO_comparison.png",
  p_GO_compare,
  width = 15,
  height = 10,
  dpi = 300
)
```

12.15 Prepare a COVID-19-associated signature for drug repurposing

```{r}
drug_signature <- covid_associated_genes |>
  filter(
    sign(covid_healthy_log2FC) ==
      sign(covid_virus_log2FC)
  ) |>
  mutate(
    direction = if_else(
      covid_virus_log2FC > 0,
      "Up",
      "Down"
    )
  )

table(
  drug_signature$direction
)

top_up <- drug_signature |>
  filter(
    direction == "Up"
  ) |>
  arrange(
    covid_virus_padj
  ) |>
  slice_head(
    n = 100
  )

top_down <- drug_signature |>
  filter(
    direction == "Down"
  ) |>
  arrange(
    covid_virus_padj
  ) |>
  slice_head(
    n = 100
  )

cat(
  "Top upregulated genes:",
  nrow(top_up),
  "\n"
)

cat(
  "Top downregulated genes:",
  nrow(top_down),
  "\n"
)
```

12.16 Save drug-repurposing input signatures

```{r}
write.table(
  top_up$gene,
  "results/signatures/COVID19_Top100_UP.txt",
  row.names = FALSE,
  col.names = FALSE,
  quote = FALSE
)

write.table(
  top_down$gene,
  "results/signatures/COVID19_Top100_DOWN.txt",
  row.names = FALSE,
  col.names = FALSE,
  quote = FALSE
)

file.exists(
  "results/signatures/COVID19_Top100_UP.txt"
)

file.exists(
  "results/signatures/COVID19_Top100_DOWN.txt"
)
```
