---
title: "Untitlgyvuged"
output: html_document
---

1. Dataset preparations
```{r}
library(GEOquery)
library(tidyverse)
library(ComplexHeatmap)
library(factoextra)
library(ggrepel)
library(GGally)
```
1.1 Load GEO data
```{r}
gse <- getGEO("GSE163151")
```
1.2 Extract metadata
```{r}
metadata <- pData(gse$GSE163151_series_matrix.txt.gz)
```
1.3 Check metadata structure
```{r}
dim(metadata)
colnames(metadata)
head(metadata)
```
1.4 Check sample titles
```{r}
head(metadata$title, 20)
```
1.5 Check sample type
```{r}
table(metadata$source_name_ch1)
```
1.6 Find all characteristics columns
```{r}
char_cols <- grep(
  "characteristics_ch1",
  colnames(metadata),
  value = TRUE
)

char_cols
```
1.7 Print unique values in each characteristics column
```{r}
for(col in char_cols){
  cat("\n---", col, "---\n")
  print(unique(metadata[[col]]))
}
```
2. Metadata preparation
```{r}
metadata_clean <- metadata |>
  transmute(
    geo_accession = geo_accession,
    sample_name = title,
    sample_type = source_name_ch1,

    disease_group = str_remove(
      characteristics_ch1.1,
      "^disease state: "
    ),

    pathogen = str_remove(
      characteristics_ch1.2,
      "^pathogen: "
    ),

    severity = str_remove(
      characteristics_ch1.4,
      "^disease severity: "
    )
  ) |>
  separate(
    sample_name,
    into = c("sample_prefix", "sample_number"),
    sep = "-",
    remove = FALSE
  )

head(metadata_clean)
```
2.1 Check separated sample names
```{r}
head(
  metadata_clean |>
    select(
      sample_name,
      sample_prefix,
      sample_number
    )
)
```
2.2 Create analysis groups
```{r}
metadata_np <- metadata_clean |>
  filter(sample_type == "NP swab") |>
  mutate(
    analysis_group = case_when(
      disease_group == "COVID-19" ~ "COVID-19",
      disease_group == "Viral acute respiratory illness" ~ "Other Virus",
      disease_group == "Donor control" ~ "Healthy",
      disease_group == "Non-viral acute respiratory illness" ~ "Nonviral",
      TRUE ~ "Exclude"
    )
  )

table(metadata_np$analysis_group)

table(
  metadata_np$analysis_group,
  metadata_np$pathogen
)
```
3. Main comparison metadata
```{r}
metadata_main <- metadata_np |>
  filter(
    analysis_group %in% c("COVID-19", "Other Virus")
  )

table(metadata_main$analysis_group)
```
3.1 Check main comparison metadata
```{r}
head(metadata_main)

table(
  metadata_main$analysis_group,
  metadata_main$pathogen
)
```
4. Download supplementary files
```{r}
getGEOSuppFiles(
  "GSE163151",
  makeDirectory = TRUE
)
```
4.1 Check downloaded files
```{r}
list.files(
  "GSE163151",
  recursive = TRUE,
  full.names = TRUE
)
```
4.2 Check tar files
```{r}
tar_files <- list.files(
  "GSE163151",
  pattern = "\\.tar$",
  recursive = TRUE,
  full.names = TRUE
)

tar_files
```
4.3 Extract tar files
```{r}
for(f in tar_files){
  untar(
    f,
    exdir = "GSE163151/samples"
  )
}
```
4.4 Find count files
```{r}
count_files <- list.files(
  "GSE163151",
  pattern = "_counts\\.txt\\.gz$",
  recursive = TRUE,
  full.names = TRUE
)

head(count_files, 20)
length(count_files)
```
4.5 Read count files
```{r}
counts_list <- count_files |>
  set_names(
    basename(count_files) |>
      str_remove("^GSM[0-9]+_") |>
      str_remove("_counts\\.txt\\.gz$")
  ) |>
  map(
    ~ read.delim(
      .x,
      header = TRUE,
      row.names = 1,
      check.names = FALSE
    )
  )
```
4.6 Check count files
```{r}
ncols <- map_int(counts_list, ncol)

table(ncols)

names(counts_list)[ncols != 1]
```
4.7 Remove invalid count file
```{r}
counts_list <- counts_list[ncols == 1]

length(counts_list)
```
4.8 Combine count files
```{r}
counts <- do.call(
  cbind,
  counts_list
)

colnames(counts) <- names(counts_list)

dim(counts)
head(counts)
```
4.9 Check sample matching
```{r}
table(
  metadata_main$sample_name %in% colnames(counts)
)
```
4.10 Keep main comparison samples
```{r}
metadata_main <- metadata_main |>
  filter(
    sample_name %in% colnames(counts)
  )

counts_main <- counts[
  ,
  metadata_main$sample_name,
  drop = FALSE
]
```
4.11 Make order of samples match
```{r}
all(
  metadata_main$sample_name ==
    colnames(counts_main)
)
```
4.12 Check final data
```{r}
dim(metadata_main)

dim(counts_main)

head(metadata_main)

head(counts_main)
```
4.13 Check raw counts
```{r}
all(
  counts_main == round(counts_main)
)

all(
  counts_main >= 0
)
```
5. Filter low-expression genes
```{r}
keep <- rowSums(counts_main >= 10) >= 10

table(keep)

counts_filtered <- counts_main[keep, ]

dim(counts_filtered)
```
5.1 Check library size
```{r}
library_size <- colSums(counts_filtered)

summary(library_size)
```
5.2 Add library size to metadata
```{r}
metadata_main$library_size <- library_size

head(metadata_main)
```
5.3 Plot library size
```{r}
ggplot(
  metadata_main,
  aes(
    x = analysis_group,
    y = library_size
  )
) +
  geom_boxplot() +
  geom_jitter(
    width = 0.2,
    alpha = 0.5
  ) +
  theme_classic() +
  labs(
    x = "Group",
    y = "Library size"
  )
```
5.4 Check number of detected genes per sample
```{r}
detected_genes <- colSums(counts_filtered > 0)

metadata_main$detected_genes <- detected_genes

summary(detected_genes)
```
5.5 Plot detected genes
```{r}
ggplot(
  metadata_main,
  aes(
    x = analysis_group,
    y = detected_genes
  )
) +
  geom_boxplot() +
  geom_jitter(
    width = 0.2,
    alpha = 0.5
  ) +
  theme_classic() +
  labs(
    x = "Group",
    y = "Detected genes"
  )
```
6. DESeq2 preparation
```{r}
library(DESeq2)
```
6.1 Set row names of metadata
```{r}
rownames(metadata_main) <- metadata_main$sample_name

all(
  rownames(metadata_main) ==
    colnames(counts_filtered)
)
```
6.2 Set analysis group
```{r}
metadata_main$analysis_group <- factor(
  metadata_main$analysis_group,
  levels = c("Other Virus", "COVID-19")
)

table(metadata_main$analysis_group)
```
6.3 Create DESeq2 dataset
```{r}
dds <- DESeqDataSetFromMatrix(
  countData = counts_filtered,
  colData = metadata_main,
  design = ~ analysis_group
)

dds
```
6.4 Normalize counts
```{r}
dds <- estimateSizeFactors(dds)

normalized_counts <- counts(
  dds,
  normalized = TRUE
)

dim(normalized_counts)
head(normalized_counts)
```
6.5 Variance stabilizing transformation
```{r}
vsd <- vst(
  dds,
  blind = TRUE
)

vsd
```
6.6 PCA
```{r}
pca_data <- plotPCA(
  vsd,
  intgroup = "analysis_group",
  returnData = TRUE
)

percent_var <- round(
  100 * attr(pca_data, "percentVar")
)
```
6.7 Plot PCA
```{r}
ggplot(
  pca_data,
  aes(
    x = PC1,
    y = PC2,
    color = analysis_group
  )
) +
  geom_point(
    size = 3,
    alpha = 0.7
  ) +
  xlab(
    paste0(
      "PC1: ",
      percent_var[1],
      "% variance"
    )
  ) +
  ylab(
    paste0(
      "PC2: ",
      percent_var[2],
      "% variance"
    )
  ) +
  theme_classic()
```
6.8 PCA with sample labels
```{r}
ggplot(
  pca_data,
  aes(
    x = PC1,
    y = PC2,
    color = analysis_group,
    label = name
  )
) +
  geom_point(
    size = 3,
    alpha = 0.7
  ) +
  geom_text_repel(
    size = 2,
    max.overlaps = 20
  ) +
  xlab(
    paste0(
      "PC1: ",
      percent_var[1],
      "% variance"
    )
  ) +
  ylab(
    paste0(
      "PC2: ",
      percent_var[2],
      "% variance"
    )
  ) +
  theme_classic()
```
6.9 Check low-depth samples
```{r}
metadata_main |>
  arrange(library_size) |>
  select(
    sample_name,
    analysis_group,
    pathogen,
    library_size,
    detected_genes
  ) |>
  head(20)
```
6.10 Check highest-depth samples
```{r}
metadata_main |>
  arrange(desc(library_size)) |>
  select(
    sample_name,
    analysis_group,
    pathogen,
    library_size,
    detected_genes
  ) |>
  head(20)
```
6.11 Sample correlation
```{r}
sample_cor <- cor(
  assay(vsd)
)

dim(sample_cor)
```
6.12 Sample correlation heatmap
```{r}
Heatmap(
  sample_cor,
  name = "Correlation",
  show_row_names = FALSE,
  show_column_names = FALSE
)
```
6.13 Find most variable genes
```{r}
gene_variance <- apply(
  assay(vsd),
  1,
  var
)

top_genes <- names(
  sort(
    gene_variance,
    decreasing = TRUE
  )
)[1:50]
```
6.14 Heatmap of most variable genes
```{r}
Heatmap(
  assay(vsd)[top_genes, ],
  name = "Expression",
  show_column_names = FALSE,
  cluster_rows = TRUE,
  cluster_columns = TRUE
)
```
6.15 Prepare COVID-19 vs Healthy metadata
```{r}
metadata_covid_healthy <- metadata_np |>
  filter(
    analysis_group %in% c(
      "COVID-19",
      "Healthy"
    )
  )
```
6.16 Check COVID-19 vs Healthy groups
```{r}
table(
  metadata_covid_healthy$analysis_group
)
```
6.17 Keep matching count samples
```{r}
metadata_covid_healthy <- metadata_covid_healthy |>
  filter(
    sample_name %in% colnames(counts)
  )

counts_covid_healthy <- counts[
  ,
  metadata_covid_healthy$sample_name,
  drop = FALSE
]
```
6.18 Check sample order
```{r}
all(
  metadata_covid_healthy$sample_name ==
    colnames(counts_covid_healthy)
)
```
6.19 Filter low-expression genes
```{r}
keep_covid_healthy <- rowSums(
  counts_covid_healthy >= 10
) >= 10

counts_covid_healthy_filtered <-
  counts_covid_healthy[
    keep_covid_healthy,
  ]

dim(
  counts_covid_healthy_filtered
)
```
7.0 Match metadata rownames with count columns
```{r}
rownames(metadata_covid_healthy) <- metadata_covid_healthy$sample_name

all(
  rownames(metadata_covid_healthy) ==
    colnames(counts_covid_healthy_filtered)
)
```
7.1 Create DESeq2 dataset
```{r}
dds_covid_healthy <- DESeqDataSetFromMatrix(
  countData = counts_covid_healthy_filtered,
  colData = metadata_covid_healthy,
  design = ~ analysis_group
)

dds_covid_healthy
```
7.2 Run DESeq2
```{r}
dds_covid_healthy <- DESeq(
  dds_covid_healthy
)
```
7.3 Extract DEG results
```{r}
res_covid_healthy <- results(
  dds_covid_healthy,
  contrast = c(
    "analysis_group",
    "COVID-19",
    "Healthy"
  )
)

res_covid_healthy
```
7.4 Convert results to dataframe
```{r}
deg_covid_healthy <- as.data.frame(
  res_covid_healthy
) |>
  rownames_to_column("gene")

head(deg_covid_healthy)
```
7.5 Remove NA values
```{r}
deg_covid_healthy <- deg_covid_healthy |>
  filter(
    !is.na(padj)
  )

dim(deg_covid_healthy)
```
7.6 Significant DEGs
```{r}
deg_sig <- deg_covid_healthy |>
  filter(
    padj < 0.05,
    abs(log2FoldChange) >= 1
  )

dim(deg_sig)
```
7.7 Count upregulated and downregulated genes
```{r}
deg_sig |>
  mutate(
    direction = case_when(
      log2FoldChange >= 1 ~ "Up",
      log2FoldChange <= -1 ~ "Down"
    )
  ) |>
  dplyr::count(direction)
```
7.8 Top DEGs
```{r}
deg_sig |>
  arrange(padj) |>
  select(
    gene,
    baseMean,
    log2FoldChange,
    pvalue,
    padj
  ) |>
  head(20)
```
7.9 MA plot
```{r}
plotMA(
  res_covid_healthy,
  ylim = c(-5, 5)
)
```
7.10 Prepare volcano plot data
```{r}
deg_covid_healthy <- deg_covid_healthy |>
  mutate(
    significance = case_when(
      padj < 0.05 & log2FoldChange >= 1 ~ "Up",
      padj < 0.05 & log2FoldChange <= -1 ~ "Down",
      TRUE ~ "Not significant"
    )
  )
```
7.11 Volcano plot
```{r}
ggplot(
  deg_covid_healthy,
  aes(
    x = log2FoldChange,
    y = -log10(padj),
    color = significance
  )
) +
  geom_point(
    alpha = 0.6,
    size = 1.5
  ) +
  geom_vline(
    xintercept = c(-1, 1),
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed"
  ) +
  theme_classic() +
  labs(
    x = "log2 Fold Change",
    y = "-log10 adjusted p-value",
    color = "Significance"
  )
```

7.12 Select top DEGs for heatmap
```{r}
top_deg_genes <- deg_sig |>
  arrange(padj) |>
  slice_head(n = 50) |>
  pull(gene)
```
7.13 VST for COVID-19 vs Healthy
```{r}
vsd_covid_healthy <- vst(
  dds_covid_healthy,
  blind = FALSE
)
```
7.14 DEG heatmap
```{r}
Heatmap(
  assay(vsd_covid_healthy)[top_deg_genes, ],
  name = "Expression",
  show_column_names = FALSE,
  cluster_rows = TRUE,
  cluster_columns = TRUE
)
```
